On the formation of H$\alpha$ line emission around classical T Tauri stars

### 2006-04-05 TTau_Halpha/revision01/kurosawa.tex
Minor typos fixed after proof reading (RK)
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@697 25464f21-6099-402c-b30a-640f1d40d38b

We now examine the sensitivity of the model profiles to the wind
collimation or the source displacement parameter $d$. As we can see
from Fig.~\ref{fig:config_discwind}, the disc wind
geometry/collimation is determined by $d$ and the wind launching radii
($R_{\mathrm{wi}}$ and $R_{\mathrm{wo}}$). Here, we fix the values of
$R_{\mathrm{wi}}$ and $R_{\mathrm{wo}}$ as in
Table~\ref{tab:std_discwind_parameters}, and inspect the model
dependency on $d$ alone.  The minimum wind opening angle (measured
from the symmetry axis) is given by
$\theta_{\mathrm{open}}=\arctan\left( R_{\mathrm{wi}}/d \right)$.
Fig.~\ref{fig:discwind_d_effect} shows the profiles computed for $d
= 2~R_{*}$, $7~R_{*}$ and $22~R_{*}$ for the inclination angle
$i=55^{\circ}$.  As the wind becomes more collimated (larger $d$), the
wind launching angle becomes smaller and the wind density is enhanced
towards the symmetry axis ($z$); hence, the wind emission increases.
When the collimation of the wind is small (as in the $d = 2~R_{*}$
case), the P-Cygni absorption is absent from the profile because the
opening angle of the disc wind is greater than the inclination angle.
For the disc wind models presented in
Fig.~\ref{fig:atlas_discwind_only}, we use $d=22~R_{*}$ 
in order to have a moderate amount of wind absorption at a moderate
inclination angle ($i\sim60^\circ$).

@@

First, we want to simulate a more \emph{colliminated} wind (toward
the symmetry axis).  Since the stream lines in our simple disc
wind are represented by straight lines, the wind never bends
toward the symmetry axis ($z$-axis) as the distance from the
disc increases unlike the model of \citet{blandford:1982}. We
therefore used a larger launching angle to simulate 
the stream lines in the far field (at a larger distance from
the disc plane) in their model.  This is an intrinsic problem
with approximating the stream lines with straight lines.

Second, the larger source displacement model would provide a greater
consistency with the fraction of H$\alpha$ profiles affected by wind
absorption seen in observations.  According to the observations of
\citet{reipurth:1996} and \citet{alencar:2000}, more than 50 per cent
of CTTS show wind absorption features in their H$\alpha$
profiles. In our model, if the launching angle is too small
($\theta_{\mathrm{L}}<60^{\circ}$), the wind absorption feature would not be seen or
would be very small for profiles computed with $i<60^{\circ}$, as
demonstrated in Fig.~~\ref{fig:discwind_d_effect}.  The situation will
however change if the field/stream line is not straight and bending
towards the symmetry axis as the wind travel higher above the disc.

@@

An examination of the profiles in
Figs.~\ref{fig:atlas_discwind_only} and \ref{fig:atlas_hybrid} reveals
some problems 
with uniqueness that must inevitably occur when trying to assign an
individual profile as wind or accretion dominated i.e. very similar
profiles may arise from very different circumstellar geometries. For
example, the disc-wind only model with
$\left(T_{\mathrm{wind}},\,\beta\right)=\left(8000\,\mathrm{K},\,1.0\right)$
in Figures~\ref{fig:atlas_discwind_only} and the accretion dominated
model with
$\left(T_{\mathrm{wind}},\,\beta\right)=\left(9000\,\mathrm{K},\,0.5\right)$
in Fig.~\ref{fig:atlas_hybrid} give very similar
profiles.

@@

\textbf{Type~III-B}:
Profiles with a blue secondary peak that is less than half the
strength of the primary peak are classified as Type~III-B. This type
of profile is the most common in the Reipurth sample, comprising 14
out of 43 (33 per cent) CTTS.
We find Type~III-B profiles only in our $\dot{M}_{{\rm acc}} = 10^{-7}\,
\mathrm{M_{\sun}\,yr^{-1}}$ models. A fast wind acceleration and
a moderate inclination are necessary, or a slower accelerating wind
viewed pole-on.  We do not see this profile type in our high
inclination models.
Although Type~III-B profiles do not occur at
$\dot{M}_{\mathrm{acc}} =
1.0\times10^{-8}\,\mathrm{M_{\sun}\,yr^{-1}}$, we have examined
the intermediate case of
$\dot{M}_{\mathrm{acc}} =3\times10^{-8}\,\mathrm{M_{\sun}\,yr^{-1}}$, which may
represent a typical T Tauri mass accretion rate
(\citealt{gullbring:1998}; \citealt{calvet:2004}). We find that
the models with fast accelerating, high temperature winds do
display Type~III-B profiles. It is clear that our model, as it
stands, requires above-average (but perhaps not atypical)
mass-accretion rates in order to produce Type~III-B profiles. It
is less clear whether the 14 Type~III-B objects found by
\citet{reipurth:1996} constitute the highest mass-accretion rate
objects in their sample: the average EW for T~Tauri stars with
Type~III-B profiles is 69\AA, which is in the top quartile of
their complete T~Tauri star sample, perhaps indicating that
Type~III-B profiles are associated with high EWs, and therefore
strong mass-accretion.

@@

\bibitem[{{Frank} {et~al.}(2002){Frank}, {King} \& {Raine}}]{frank:2002}
{Frank} J., {King} A., {Raine} D.~J., 2002, {Accretion Power in Astrophysics,
  3rd edn.} Cambridge Univ.\, Press, Cambridge, p. 398

@@

\bibitem[{Grevesse \& Noels(1993)}]{grevesse:1993}
Grevesse N., Noels A., 1993, in Origin and Evolution of the Elements, N.\ P.,
  E.\ V.-F., M.\ C., eds., Cambridge Univ.\, Press, Cambridge, p.~15

@@

\bibitem[{{Pudritz} \& {Banerjee}(2005)}]{pudritz:2005}
{Pudritz} R.~E., {Banerjee} R., 2005, in Proc.~IAU Symp. Massive star
birth: A crossroads of Astrophysics, Cesaroni, R., Felli, M.,
Churchwell, E., Walmsley, M., eds, Cambridge Univ.\, Press, Cambridge,
225, p.~163

>>>>

We now examine the sensitivity of the model profiles to the wind
collimation or the source displacement parameter $d$. As we can see
from Fig.~\ref{fig:config_discwind}, the disc wind
geometry/collimation is determined by $d$ and the wind launching radii
($R_{\mathrm{wi}}$ and $R_{\mathrm{wo}}$). Here, we fix the values of
$R_{\mathrm{wi}}$ and $R_{\mathrm{wo}}$ as in
Table~\ref{tab:std_discwind_parameters}, and inspect the model
dependency on $d$ alone.  The minimum wind opening angle (measured
from the symmetry axis) is given by
$\theta_{\mathrm{open}}=\arctan\left( R_{\mathrm{wi}}/d \right)$.
Fig.~\ref{fig:discwind_d_effect} shows the profiles computed for $d
= 2~R_{*}$, $7~R_{*}$ and $22~R_{*}$ for the inclination angle
$i=55^{\circ}$.  As the wind becomes more collimated (larger $d$), the
wind opening angle becomes smaller and the wind density is enhanced
towards the symmetry axis ($z$); hence, the wind emission increases.
When the collimation of the wind is small (as in the $d = 2~R_{*}$
case), the P-Cygni absorption is absent from the profile because the
opening angle of the disc wind is greater than the inclination angle.
For the disc wind models presented in
Fig.~\ref{fig:atlas_discwind_only}, we use $d=22~R_{*}$ 
in order to have a moderate amount of wind absorption at a moderate
inclination angle ($i\sim60^\circ$).

@@

First, we want to simulate a more \emph{collimated} wind (toward
the symmetry axis).  Since the stream lines in our simple disc
wind are represented by straight lines, the wind never bends
toward the symmetry axis ($z$-axis) as the distance from the
disc increases unlike the model of \citet{blandford:1982}. We
therefore used a larger launching angle to simulate 
the stream lines in the far field (at a larger distance from
the disc plane) in their model.  This is an intrinsic problem
with approximating the stream lines with straight lines.

Second, the larger source displacement model would provide a greater
consistency with the fraction of H$\alpha$ profiles affected by wind
absorption seen in observations.  According to the observations of
\citet{reipurth:1996} and \citet{alencar:2000}, more than 50 per cent
of CTTS show wind absorption features in their H$\alpha$
profiles. In our model, if the launching angle is too small
($\theta_{\mathrm{L}}<60^{\circ}$), the wind absorption feature would not be seen or
would be very small for profiles computed with $i<60^{\circ}$, as
demonstrated in Fig.~\ref{fig:discwind_d_effect}.  The situation will
however change if the field/stream line is not straight and bending
towards the symmetry axis as the wind travel higher above the disc.

@@

An examination of the profiles in
Figs.~\ref{fig:atlas_discwind_only} and \ref{fig:atlas_hybrid} reveals
some problems 
with uniqueness that must inevitably occur when trying to assign an
individual profile as wind or accretion dominated i.e. very similar
profiles may arise from very different circumstellar geometries. For
example, the disc-wind only model with
$\left(T_{\mathrm{wind}},\,\beta\right)=\left(8000\,\mathrm{K},\,1.0\right)$
in Fig.~\ref{fig:atlas_discwind_only} and the accretion dominated
model with
$\left(T_{\mathrm{wind}},\,\beta\right)=\left(9000\,\mathrm{K},\,0.5\right)$
in Fig.~\ref{fig:atlas_hybrid} give very similar
profiles.

@@

\textbf{Type~III-B}:
Profiles with a blue secondary peak that is less than half the
strength of the primary peak are classified as Type~III-B. This type
of profile is the most common in the Reipurth sample, comprising 14
out of 43 (33 per cent) CTTS.
We find Type~III-B profiles only in our $\dot{M}_{{\rm acc}} = 10^{-7}\,
\mathrm{M_{\sun}\,yr^{-1}}$ models. A fast wind acceleration and
a moderate inclination are necessary, or a slower accelerating wind
viewed pole-on.  We do not see this profile type in our high
inclination models.
Although Type~III-B profiles do not occur at
$\dot{M}_{\mathrm{acc}} =
1.0\times10^{-8}\,\mathrm{M_{\sun}\,yr^{-1}}$, we have examined
the intermediate case of
$\dot{M}_{\mathrm{acc}} =3\times10^{-8}\,\mathrm{M_{\sun}\,yr^{-1}}$, which may
represent a typical T Tauri mass accretion rate
(\citealt{gullbring:1998}; \citealt{calvet:2004}). We find that
the models with fast accelerating, high temperature winds do
display Type~III-B profiles. It is clear that our model, as it
stands, requires above-average (but perhaps not atypical)
mass-accretion rates in order to produce Type~III-B profiles. It
is less clear whether the 14 Type~III-B objects found by
\citet{reipurth:1996} constitute the highest mass-accretion rate
objects in their sample: the average EW for T~Tauri stars with
Type~III-B profiles is 69~\AA, which is in the top quartile of
their complete T~Tauri star sample, perhaps indicating that
Type~III-B profiles are associated with high EWs, and therefore
strong mass-accretion.

@@

\bibitem[{{Frank} {et~al.}(2002){Frank}, {King} \& {Raine}}]{frank:2002}
{Frank} J., {King} A., {Raine} D.~J., 2002, {Accretion Power in Astrophysics,
  3rd edn.} Cambridge Univ.\, Press, Cambridge, p.~398

@@

\bibitem[{Grevesse \& Noels(1993)}]{grevesse:1993}
Grevesse N., Noels A., 1993, in {Prantzos} N., {Vangioni-Flam} E., {Casse}
M., eds,  Origin and Evolution of the Elements,  Cambridge Univ.\, Press, Cambridge, p.~15

@@

\bibitem[{{Pudritz} \& {Banerjee}(2005)}]{pudritz:2005}
{Pudritz} R.~E., {Banerjee} R., 2005, in {Cesaroni}, R., {Felli}, M.,
{Churchwell}, E., {Walmsley}, M., eds, Proc.~IAU Symp.~225, Massive star
birth: A crossroads of Astrophysics,  Cambridge Univ.\, Press,
Cambridge, p.~163
